Dual Targeting of Histone Deacetylases and MYC as Potential Treatment Strategy for H3-K27M Pediatric Gliomas

Version 1

Abstract

Abstract

Diffuse midline gliomas (DMG) are aggressive and fatal pediatric tumors of the central nervous system that are highly resistant to treatments. Lysine to methionine substitution of residue 27 on histone H3 (H3-K27M) is a driver mutation in DMGs, reshaping the epigenetic landscape of these cells to promote tumorigenesis. H3-K27M gliomas are characterized by deregulation of histone acetylation and methylation pathways, as well as the oncogenic MYC pathway. In search of effective treatment, we examined the therapeutic potential of dual targeting of histone deacetylases (HDACs) and MYC in these tumors. Treatment of H3-K27M patient-derived cells with Sulfopin, an inhibitor shown to block MYC-driven tumors in-vivo, in combination with the HDAC inhibitor Vorinostat, resulted in substantial decrease in cell viability. Moreover, transcriptome and epigenome profiling revealed synergistic effect of this drug combination in downregulation of prominent oncogenic pathways such as mTOR. Finally, in-vivo studies of patient-derived orthotopic xenograft models showed significant tumor growth reduction in mice treated with the drug combination. These results highlight the combined treatment with PIN1 and HDAC inhibitors as a promising therapeutic approach for these aggressive tumors.

Introduction

Diffuse midline gliomas (DMGs) are fatal tumors of the central nervous system that occur primarily in children1,2. Years of clinical trials have revealed that conventional chemotherapy is ineffective, and with no option for complete resection, DMG is now the leading cause of brain tumor-related death in children3,4. Sequencing studies showed that up to 80% of the diagnosed children exhibit a lysine 27 to methionine (H3-K27M) mutation in one of the genes encoding histone H3.3 and H3.1 (clinically classified as DMG, H3 K27M–mutant). Numerous subsequent studies revealed pronounced alterations in the epigenetic landscape driven by this histone mutation5–8. Among these global epigenetic alterations are a drastic loss of H3 lysine 27 tri-methylation (H3K27me3), concomitant with an increase in H3 lysine 27 acetylation (H3K27ac). These epigenetic alterations were shown to affect transcriptional programs and support tumorigenesis 9–14.

Detailed characterization of the modes of chromatin deregulation in these tumors pointed toward targeting epigenetic pathways as a promising therapeutic approach15. Specifically, pervasive H3K27 acetylation patterns observed in H3-K27M DMGs, along with enrichment of H3K27ac specifically on mutant nucleosomes11–13,16, underscored the histone acetylation pathway as a potential drug target for this cancer. Indeed, pan-histone deacetylase inhibitors (HDACi), such as Vorinostat and Panobinostat, were found to inhibit growth and to restore gene expression alterations observed in H3-K27M malignant gliomas17–19. However, recent studies demonstrated that H3-K27M cells develop resistance to HDACi treatment17,19, stressing the importance of identifying novel drug combinations to induce synergistic inhibition of oncogenic pathways.

The MYC oncogenic pathway is frequently deregulated in diverse cancers, and thus stands as a prominent therapeutic target20. High-level gene amplification of MYC is recurrently observed in pediatric high-grade gliomas21,22. H3-K27M gliomas show high expression of MYC and MYC target genes, due to both epigenetic alterations and structural variants, resulting in a viability dependency on MYC signaling12,23,24. Unfortunately, there are currently no clinical inhibitors to directly target the MYC protein25. However, Dubiella et al. have recently developed a unique inhibitor targeting Peptidyl-prolyl isomerase NIMA-interacting-1 (PIN1), an upstream regulator of MYC activation26. PIN1 itself is frequently upregulated in many types of cancers, resulting in sustained proliferative signaling and tumor growth27,28. The PIN1 inhibitor, Sulfopin, was shown to downregulate MYC target genes in-vitro and to block MYC-driven tumors in murine models of neuroblastoma26.

To address the increasing need for an effective combination therapy for H3-K27M DMGs, we tested whether dual targeting of histone acetylation and MYC activation, using Vorinostat and Sulfopin, would yield beneficial outcome. We show that the combined treatment substantially reduces viability of patient-derived glioma cells, in concordance with inhibition of prominent oncogenic pathways, including the mTOR pathway. Treatment of H3-K27M DMG xenograft mice models with this drug combination led to reduced tumor growth and confirmed downregulation of mTOR in-vivo. Together, these results suggest the combination of PIN1 and HDAC inhibitors as a promising therapeutic strategy for H3-K27M gliomas.

Results

Sulfopin inhibits MYC targets and reduces viability of H3-K27M glioma cells

In light of the important role of MYC in H3-K27M DMGs, we set to explore whether the PIN1 inhibitor, Sulfopin, may affect MYC signaling in these cells and thus confer therapeutic potential. Sulfopin dependent inhibition of PIN1 was shown to downregulate the expression of MYC target genes in several cancer cell lines26. Indeed, RNA sequencing analysis of patient-derived glioma cells (SU-DIPG13) treated with Sulfopin revealed strong enrichment of MYC targets among the downregulated genes (Fig. 1A). The genes downregulated by Sulfopin were also enriched for MYC-bound genes identified in several non-glioma cell lines, supporting a role for PIN1 in activation of the MYC pathway also in these glioma cells (Fig. S1A-B). Of note, Sulfopin was developed as a covalent inhibitor of PIN1, and did not lead to PIN1 degradation, as previously shown (Fig. S1C) 26. We further validated the decreased expression of prominent MYC targets27,29,30 upon Sulfopin treatment by RT-qPCR analysis (Fig. 1B). Interestingly, Cut&Run analysis31 revealed an increase in H3K27me3 levels, indicative of repressed chromatin32, around the transcriptional start sites (TSS) of MYC target genes, confirming epigenetic silencing of these genes upon PIN1 inhibition (Fig. 1C, S1D). Of note, global H3K27me3 distribution on all TSS and gene-body regions was not affected by Sulfopin treatment, supporting a specific increase in H3K27me3 on the TSS of MYC target genes (Fig. S1E-F). Furthermore, Sulfopin-unique H3K27me3 peaks (i.e., peaks that were found only in the Sulfopin-treated cells, and not in DMSO-treated cells) were enriched for MYC target genes (Fig. 1D).

Sulfopin inhibits MYC signaling and reduces viability of DMG cells in a H3-K27M-dependent manner

(A) Functional enrichment analysis on significantly downregulated genes in SU-DIPG13 cells treated with 10uM Sulfopin for 12 hours, compared to DMSO. Enrichr algorithm123 was used to compare downregulated genes against the Molecular Signatures Database (MSigDB) hallmark geneset119. Dashed line denotes adjusted p-value = 0.05. MYC targets are significantly enriched among the Sulfopin downregulated genes.

(B) RT-qPCR analysis of selected MYC target genes in SU-DIPG13 cells treated with 10uM Sulfopin for 12 hours, compared to DMSO. Fold change between Sulfopin and DMSO treated cells was calculated and the mean ± SD of two technical repeats is shown.

(C) H3K27me3 Cut&Run read coverage over MYC target genes (‘MYC Targets V1’ hallmark geneset119, n=200), in SU-DIPG13 cells treated with 10uM Sulfopin for 8 days compared to DMSO. Sulfopin treatment increases H3K27me3 levels on the TSS of MYC target genes.

(D) Functional enrichment analysis of the genes associated with Sulfopin-unique H3K27me3 peaks, in SU-DIPG13 cells treated as in C. Dashed line denotes adjusted p-value = 0.05. MYC target geneset (‘MYC Targets V1’ hallmark geneset119) is mildly enriched among these genes, with adjusted p-value of 0.077.

(E) Cell viability, as measured by CellTiterGlo, of eight DMG cultures (H3.3K27M: SU-DIPG13, SU-DIPG6, SU-DIPG17, SU-DIPG25 and SU-DIPG50. H3.1K27M: SU-DIPG36, SU-DIPG38 and SU-DIPG21), treated with Sulfopin for eight days with pulse at day four, compared to DMSO. Mean±SD of two technical replicates is shown. Logarithmic scale is used for the x-axis. Sulfopin treatment led to a mild reduction in cell viability in all H3-K27M glioma cultures.

(F) Cell viability, as measured by CellTiterGlo, of two isogenic DMG cell lines (SU-DIPG13 and BT245) in which the mutant histone was knocked-out (KO), treated with the indicated concentration of Sulfopin for 8 days, compared to DMSO. For each cell line and concentration, the fold change in viability between Sulfopin and DMSO treated cells is shown. For SU-DIPG13-mean ± SE of at least two independent experiments is shown. For BT245-mean± SD of three technical replicates is shown. H3-K27M glioma cells show higher sensitivity to Sulfopin treatment compared to the KO cells. *P < 0.05; **P < 0.01 (two-sample t-test over all technical replicates). Significance adjusted after Bonferroni correction.

We next explored the effect of Sulfopin on the viability of eight patient derived DMG cultures, harboring either the H3.3-K27M or H3.1-K27M mutation. In line with downregulation of MYC signaling, Sulfopin treatment led to a mild reduction in cell viability, conserved across the different cultures (Fig. 1E). To examine whether H3-K27M mutation renders DMG cells more sensitive to Sulfopin treatment, we took advantage of two sets of isogenic patient-derived DMG cell-lines (SU-DIPG13, BT245) in which the H3F3A-K27M mutant allele has been knocked-out (KO). In both lines, the reduction in cell viability was partially rescued in the KO cells (Fig. 1F), suggesting higher sensitivity of H3-K27M mutant cells to MYC inhibition by Sulfopin. These results are in line with previous reports showing H3-K27M-dependent activation of the MYC/RAS axis in mouse and human cells12,23,24. Of note, DMG cells showed greater sensitivity to the inhibitor compared to human astrocytic-like cells (Fig. S1G).

To further explore the status of MYC and PIN1 as indicators of response to Sulfopin inhibition, we analyzed published transcriptomic dataset of DMG tumors35. Interestingly, while DMG samples showed significantly higher expression of MYC and its target genes compared to normal samples, PIN1 expression was lower in the cancer compared to matched normal tissue (Fig. S1H-J). Furthermore, H3-K27M mutant tumors showed higher expression of MYC, compared to H3 WT tumors, while PIN1 levels were reduced across samples22 (Fig. S1K-L). These results suggest that MYC expression levels, and not PIN1, may determine the sensitivity of H3-K27M cells to Sulfopin treatment. To examine this notion, we measured the expression levels of MYC, as well as the doubling time of the different DMG cultures. The response to Sulfopin treatment (as reflected by IC50 measurements) correlated with the cells’ doubling time (Fig. S1M). Cells that divide faster were more sensitive to Sulfopin treatment, in line with PIN1 role in regulating cell cycle progression, and specifically G2/M transition36,37. As expected, the doubling time negatively correlated with MYC expression, suggesting that MYC levels may predict the response to Sulfopin (Fig S1N). Indeed, MYC levels negatively correlated with the IC50 of Sulfopin, thus, cells relaying on high MYC activity showed higher sensitivity to Sulfopin (Fig S1O).

H3K27me3 levels were shown to be drastically deregulated in DMG cells expressing H3-K27M, with global loss of this modification across the genome, and retention of residual H3K27me3 at CpG islands (CGIs). These residual H3K27me3 levels were shown to be necessary to silence tumor suppressor genes and maintain cell proliferation in DMG cells33,34. Interestingly, Sulfopin treatment resulted in reduction of H3K27me3 signal over CGIs, perhaps contributing to Sulfopin-dependent growth inhibition in these cells (Fig. S1P). Taken together, our results indicate that PIN1 inhibition by Sulfopin downregulated MYC target genes, reduced H3K27me3 levels at CGIs, and led to reduce viability of DMG cells in a H3-K27M-dependant manner.

Combination of Sulfopin and Vorinostat elicits robust inhibition of oncogenic pathways and an additive effect on cell viability

Inspired by the positive outcome of combinatorial drug therapy in H3-K27M DMG models17,38, we aimed to explore whether combining Sulfopin treatment with an epigenetic drug will further reduce cell viability. H3-K27M gliomas are characterized by deregulation of epigenetic pathways affecting post-translational modifications of lysine 27 (methylation and acetylation) and lysine 4 (methylation) on histone H39–12,16,39. Thus, we explored the effects of combining Sulfopin with four epigenetic inhibitors targeting critical components associated with these modifications: (1) EPZ-6438, an inhibitor of EZH2, the catalytic subunit of the PRC2 complex that deposits H3K27me340. H3-K27M cells were reported to be sensitive to EZH2 inhibition as they rely on the residual H3K27me3 levels to suppress key tumor-suppressor genes13,33,41,42; (2) GSK-J4, an inhibitor of the H3K27me3 demethylase JMJD343, shown to rescue H3K27me3 levels in H3-K27M gliomas17,44; (3) MM-102, an inhibitor of the H3K4me3 methyltransferase MLL140, shown to reduce viability of H3-K27M-mutant cells16; and (4) Vorinostat, an HDAC inhibitor that was shown to have clinical benefit in H3-K27M gliomas17–19. We found that the combination of Sulfopin and Vorinostat (HDACi) had the strongest effect, reducing viability by 80% (Fig. S2A).

Next, we adjusted treatment protocols to account for the optimal treatment period for each drug, and examined the effect of this drug combination on the viability of eight patient-derived DMG cultures, across different dosages (Fig. 2A-B and S2B-D). For each pair of concentrations, we calculated a BLISS index; an additive effect of the two drugs was determined as BLISS index ∼145,46 (Fig. 2C, S2E). While the combined treatment was effective in all the cultures, cells harboring H3.3-K27M showed higher sensitivity (Fig. 2D, S2F). Moreover, while an additive effect of Sulfopin and Vorinostat was detected in all the H3.3-K27M DMG cultures in most of the drug doses, for the H3.1-K27M cultures we measured an additive effect only in higher dosages of Vorinostat (Fig. 2C-E and S2E-F). In line with the difference in sensitivity, H3.1-and H3.3-K27M cells differed substantially in MYC expression levels, with H3.1-K27M cells expressing significantly lower levels of MYC (Fig. S2G). Indeed, BLISS indexes negatively correlated with MYC expression levels (Fig. 2F), indicating that H3.3-K27M cells, which express higher levels of MYC, are more sensitive to the combination of Sulfopin and Vorinostat. Importantly, this highlights MYC, and MYC target genes, as potential biomarkers of response for the combined treatment. For example, the expression levels of the two MYC targets, NOC4L and PSMD3, negatively correlated with the BLISS indexes and significantly differed between H3.1-and H3.3-K27M cells (Fig. 2F and S2H). Furthermore, these and additional MYC targets were downregulated following the combined treatment in the H3.3-K27M culture, SU-DIPG13 (Fig. S2I). Overall, our data demonstrates that combining Sulfopin treatment with Vorinostat results in an additive effect on the viability of DMG cultures, with H3.3-K27M cells showing higher sensitivity. This is in line with previous studies highlighting clinical and molecular distinction between H3.1-and H3.3-K27M mutant tumors, which is thought to dictate their targetable vulnerabilities5,47,48.

Combination of Sulfopin and Vorinostat elicits robust downregulation of oncogenic pathways

(A) Timeline demonstrating the treatment protocol for the combination of Sulfopin and Vorinostat.

(B) Percentage of cell viability, as measured by CellTiterGlo of SU-DIPG13 cells treated with Sulfopin and Vorinostat at the indicated concentrations, compared to DMSO.

(C) BLISS index measured as the ratio between the observed and the expected effect of the combination of Sulfopin and Vorinostat, for each pair of concentrations, in SU-DIPG13.Synergy: Bliss <1, Additive: Bliss=1, Antagonist: Bliss>1.

(D) Cell viability as measured by CellTiterGlo, of eight DMG cultures treated with Sulfopin (0uM, 2.5uM, 5uM, 10uM, 20uM and 40uM) and Vorinostat (1uM), compared to DMSO. H3.3-K27M and H3.1-K27M cultures are indicated in blue and orange, respectively. Mean±SD of two technical replicates is shown. H3.3-K27M cells showed higher sensitivity to the combined treatment compared to H3.1-K27M cells.

(E) The BLISS index of the combination of Sulfopin (10uM) and Vorinostat (1uM), in the indicated cultures. An additive effect was detected in all the H3.3-K27M cultures at this set of concentrations.

(F) Pearson correlation coefficient matrix of BLISS index of the combined treatment (Sulfopin (10uM) and Vorinostat (1uM)) and mRNA levels of MYC and its target genes, in the eight DMG cultures tested. mRNA levels were measured by RT-qPCR (Fig. S2F-G). Blue and yellow colors indicate negative or positive correlation, respectively. Negative correlation was detected between the BLISS indexes and the expression levels of MYC and its target genes.

(G) Unsupervised hierarchical clustering of expression levels of 620 significantly DE genes detected in SU-DIPG13 cells treated with either Sulfopin (10uM, 8 days), Vorinostat (1uM, 72 hours), the combination of Sulfopin and Vorinostat or DMSO. Gene expression rld values (log2 transformed and normalized) were standardized for each gene (row) across all samples. Color intensity corresponds to the standardized expression, low (blue) to high (red). Clusters 1 and 4 demonstrate additive transcriptional patterns associated with the combined treatment.

(H) Top: Gene Set Enrichment Analysis (GSEA) on SU-DIPG13 treated with combination of 10uM Sulfopin and 1uM Vorinostat compared to DMSO, showing significant downregulation of mTORC1 signaling (‘HALLMARK_MTORC1_SIGNALING’ geneset119) in the combined treatment. NES: Normalized Enrichment Score. FDR: false discovery rate. Bottom: Expression levels of significantly DE genes detected in the combined treatment compared to DMSO that are part of the mTORC1 signaling geneset. MTOR gene was added manually to the heat-map. Heatmaps were generated as described in B.

(I) Top: Gene Set Enrichment Analysis (GSEA) on SU-DIPG13 treated as in C, showing significant downregulation of the epigenetic BMI-1 pathway and the oncogenic cAMP pathway in the combined treatment (BMI1_DN.V1_UP; CAMP_UP.V1_UP; MSigDB C6 oncogenic signature120,121). Bottom: Expression levels of significantly DE genes detected in the combined treatment compared to DMSO that are part of the BMI-1 and cAMP genesets. Heatmaps were generated as in described B.

(J) Western blot of SU-DIPG13 treated either with Sulfopin (10uM, 8 days), Vorinostat (1uM, 72 hours), the combination of Sulfopin and Vorinostat or DMSO, using the indicated antibodies. β-tubulin is used as loading control.

To gain more insights on the molecular mechanisms underlying this effect on viability, we profiled the transcriptome of SU-DIPG13 cells (H3.3-K27M) treated with each of the drugs alone and the combined treatment (Fig. S2J). A total of 494 genes were significantly differentially expressed (DE) in cells treated with the drug combination, compared to only 145 and 167 in cells treated solely with Sulfopin or Vorinostat, respectively (Fig. S2K-L). Hierarchical clustering of all 620 DE genes confirmed strong additive effect of the combined treatment on gene expression, primarily demonstrated by clusters 1 and 4 (Fig. 2G). Cluster 1, consisting of approximately half of the DE genes, was strongly downregulated by the combined treatment, compared to single-agent treated cells. This cluster included genes that play crucial roles in glioma progression such as MTOR, ATF3, CREB3 and TGFA49–52 (Fig. S2M). Moreover, the mTORC1 signaling pathway was highly enriched within the genes comprising cluster 1 (Fig. S2N). Importantly, GSEA analysis revealed that this pathway was significantly enriched in downregulated genes following the combined treatment (Fig. 2H and S2O). Of note, additional epigenetic and oncogenic pathways such as BMI1, cAMP and IL-6/JAK/STAT3 were significantly downregulated in cells treated with the drug combination53–55 (Fig. 2I and S2O). Reduction in mTOR signaling (mTOR, phospho-mTOR and phospho-S656) and ATF3 was further validated by western blot and RT-qPCR analysis (Fig. 2J, S2P). Importantly, we also confirmed that the combined treatment with Sulfopin and Vorinostat led to induction of p21 (CDKN1A), a prominent tumor suppressor promoting cell cycle arrest and regulated by mTOR signaling57–59 (Fig. 2H, J). Finally, exploring mTOR expression in all the tested DMG cultures, revealed a strong negative correlation with the BLISS indexes of the combined treatment (Fig. S2Q), suggesting mTOR as a potential biomarker of response for the combined treatment of Sulfopin and Vorinostat. Overall, our data revealed that the combined inhibition of HDACs and MYC results in robust downregulation of prominent oncogenic pathways.

Sulfopin attenuates HDACi-mediated accumulation of H3K27ac on genomic regions associated with oncogenic pathways

We next examined epigenetic features that may underly the inhibition of oncogenic pathways by the combined treatment. Inhibition of histone deacetylases is expected to increase and alter genome-wide histone acetylation patterns, including the H3K27ac mark associated with active promoters and enhancers60. Indeed, global quantification by single-molecule measurements of H3K27-acetylated nucleosomes, as well as western blot analysis, confirmed HDACi-mediated increase in H3K27ac global levels (Fig. 3A-C, S3A). Interestingly, treatment with Sulfopin alone had the opposite effect, reducing the percentage of acetylated nucleosomes (Fig. 3B-C, S3A). As acetylation is coupled with transcription activation, this is in line with the function of Sulfopin as an inhibitor of MYC transcriptional activity26,61. As a result, the global HDACi-mediated increase in histone acetylation was also attenuated in the combined treatment with Sulfopin (Fig. 3B-C, S3A).

The combined treatment attenuates H3K27ac levels on oncogenic targets

(A) Scheme of the single-molecule imaging experimental setup133: cell-derived mono-nucleosomes are anchored in a spatially distributed manner on polyethylene glycol (PEG)-coated surface. Captured nucleosomes are incubated with fluorescently labeled antibodies directed against the H3K27ac modification. Total internal reflection fluorescence (TIRF) microscopy is utilized to record the position and modification state of each nucleosome. Time series images are taken to allow detection of maximal binding events.

(B) Single-molecule imaging quantification of the percentage of H3K27ac nucleosomes, in SU-DIPG13 cells treated with either Sulfopin (10uM, 8 days), Vorinostat (1uM, 72 hours), or the combination of Sulfopin and Vorinostat, normalized to DMSO. Mean fold ± SE of at least two independent experiments is shown. H3K27ac global levels are lower in the combined treatment compared to cells treated solely with Vorinostat. *P < 0.05 (two sample t-test).

(C) SU-DIPG13 cells were treated as in B, and analyzed by western blot using the indicated antibodies.

(D) Left panel: Heatmap shows H3K27ac read coverage around the TSS (+/-5Kb) of the significantly DE genes shown in figure 2B, in SU-DIPG13 cells treated with the combination of 10uM Sulfopin and 1uM Vorinostat versus DMSO. Average coverage is shown on top. Color intensity corresponds to the standardized expression. Clusters 1-4 are indicated. Right panel: The log2 ratio of H3K27ac read coverage in SU-DIPG13 cells treated with the combination of 10uM Sulfopin and 1uM Vorinostat vs. DMSO was calculated. Heatmap shows the ratio around the TSS (+/-5Kb) of the significantly DE genes shown in figure 2B, and average coverage is shown on top. Color intensity corresponds to the ratio between samples, low (red) to high (blue). Clusters 1-4 are indicated, with cluster 1 presenting the strongest local decrease in H3K27ac following the combined treatment compared to DMSO.

(E) IGV tracks of MTOR and SLC7A5 gene promoters, showing H3K27ac coverage in SU-DIPG13 cells treated as indicated.

(F) Functional enrichment analysis of the genes linked to enhancers (top targets of high confident enhancers) marked with H3K27ac exclusively in SU-DIPG13 cells treated with Vorinostat, and not in the combined treatment. gProfiler algorithm126 was used to calculate enrichment against the KEGG pathways DB128. Dashed line denotes adjusted p-value = 0.05. Genes associated with Vorinostat-unique enhancers are enriched for oncogenic signaling pathways.

(G) IGV track of AKT3 and SEC13 linked enhancers, showing H3K27ac coverage in SU-DIPG13 cells treated with 1uM Vorinostat or the combination of 10uM Sulfopin and 1uM Vorinostat.

(H) Normalized expression levels of AKT3 and SEC13 genes in SU-DIPG13 cells treated as in G. Mean ± SD of three technical repeats is shown.*P < 0.05 (two-sample t-test).

To profile changes in the genomic distribution of H3K27ac and H3K27me3, we applied Cut&Run in SU-DIPG13 cells treated with Vorinostat alone or in combination with Sulfopin. We first examined the promoters of genes differentially expressed in the combined treatment (as clustered in Fig. 2G). Genes comprising cluster 1 showed higher basal expression levels compared to genes from cluster 4, and these expression differences were also reflected in higher H3K27ac at these genes’ promoters (Fig. S3B-C). Moreover, in accordance with the reduction in RNA expression levels of genes comprising cluster 1, the combined treatment led to the strongest local reduction of H3K27ac levels associated with their promoters (Fig. 3D). Specifically, we found that several of the oncogenes that comprise cluster 1, such as mTOR, and mTOR regulator-SLC7A5, RELB and COPB2, showed decreased levels of H3K27ac at their promoters following the combined treatment62–64 (Fig. 3E, S3D). Cluster 1 genes also gained H3K27me3 on their promoters upon the combined treatment, supporting their epigenetic repression following Sulfopin and Vorinostat treatment (Fig. S3E).

We further identified genomic regions in which histone acetylation is lost in a Sulfopin-dependent manner (i.e., H3K27ac peaks that are present in Vorinostat-treated cells, but are lost in the combined treatment with Sulfopin). Genome distribution analysis of these unique peaks revealed that the majority of them localize to distal regions (Fig. S3F) and more than 60% of them localize to annotated enhancers. Importantly, genes associated with these enhancers were strongly enriched for oncogenic pathways such as RAS, MAPK and PI3K-AKT signaling pathways66–68 (Fig. 3F, S3G). Specifically, the combined treatment reduced H3K27ac levels on several enhancers linked to genes involved in mTOR pathway and gliomagenesis, such as: AKT3, JAK1 and NFKB2 genes, that were shown to have critical role in malignant gliomas69–73, SEC13 gene that indirectly activates mTOR74,75, and CD93, a key regulator in glioma76 (Fig 3G and S3H). Concomitant with the lower H3K27ac levels associated with these enhancers, we also observed lower expression of the associated genes (Fig. 3H and S3I). Taken together, the results show attenuation of HDACi-mediated accumulation of H3K27ac levels upon combined treatment with Sulfopin, associated with promoters and enhancers of prominent oncogenes. This is in-line with reduced expression of these oncogenic pathways, strengthening the therapeutic potential of the combined treatment.

The combined treatment reduces tumor growth in a DMG xenograft mouse model

Finally, we examined whether the combined inhibition of HDACs and PIN1 elicits an additive effect also in-vivo. Thus, we established a H3-K27M DMG orthotopic xenograft model by injections of SU-DIPG13P* cells, engineered to express firefly luciferase, to the pons of immunodeficient mice17. Mice were given ten days to develop tumors, and then randomly assigned to four groups: control mice treated with DMSO, mice treated with Sulfopin or Vorinostat as single agents, and mice treated with the combination of Sulfopin and Vorinostat. Tumor growth was monitored by in in-vivo bioluminescence imaging. In-line with our in-vitro results, we found significant reduction in tumor growth following the combined treatment compared to control DMSO treated mice (Fig. 4A-B). Importantly, the reduction was also seen when comparing to mice treated solely with Vorinostat, supporting the additive value of Sulfopin to the well-known anti-tumorigenic effect of HDACi in H3-K27M gliomas17 (Fig. 4A-B). Of note, mice from all groups showed signs of severe dehydration and general deterioration after 18 days of treatment, and thus we could not measure a survival benefit for the combined treatment.

The combined treatment with Sulfopin and Vorinostat reduces tumor growth in-vivo

(A-B) SU-DIPG13P* cells were injected to the pons of immunodeficient mice to form tumors. Ten days post injection, mice were treated for 18 days with either DMSO, Sulfopin, Vorinostat, or the combination of Sulfopin and Vorinostat. (A) In-vivo bioluminescent imaging of DMG xenografts following 18 days of treatment. The heat map superimposed over the mouse head represents the degree of photon emission by DMG cells expressing firefly luciferase. (B) DMG xenograft tumor growth as measured by change in bioluminescent photon emission following 15 days of treatment with either DMSO (n=8), Sulfopin (n=4), Vorinostat (n=5) or the combination of Sulfopin and Vorinostat (n=6). Data points represent the fold-change in maximum photon flux between day 3 and day 18 under treatment for each mouse. *P < 0.05 (two-tailed Mann-Whitney U-test).

(C-D) Immunofluorescent staining of brain sections from mice treated with DMSO (n=4) or the combination of Sulfopin and Vorinostat (n=4). (C) Representative fluorescence images of H3-K27M (red) and mTOR (green). (D) Percentage of mTOR positive cells out of the total H3-K27M-positive cells. H3-K27M positive cells show lower levels of mTOR following the combined treatment compared to DMSO. *P < 0.05 (two-tailed t-test).

To link our in-vitro transcriptomic analysis to the in-vivo effect on tumor growth, we assessed the percentage of H3-K27M cells that are positive to mTOR signal (double-positive), out of the total number of H3-K27M cells, by immunofluorescence staining. The combined treatment significantly reduced the fraction of the double-positive cells, compared to DMSO (Fig. 4C-D). Of note, mTOR signal in the mouse cells surrounding the tumor showed no significant differences between the DMSO and the combination-treated mice (Fig S4A-B). Overall, these results verify the potency of this treatment in restricting oncogenic pathways also in-vivo, and the potential clinical benefit of PIN1 and HDAC combined inhibition in H3-K27M-mutant DMGs.

Discussion

Despite decades of efforts to improve treatment for children diagnosed with H3-K27M-mutant DMGs, the standard of care remains solely radiation, with no significant improvement in overall survival1,2,79. High throughput drug screens in patient-derived cells identified HDAC inhibition as a promising therapeutic strategy, and ongoing clinical trials revealed preliminary evidence of clinical benefit38,80 (NCT02717455, NCT03566199). However, emerging resistance to HDACi in glioma cells17, as well as limited therapeutic benefits of HDACi treatment as a sole agent81,82, highlight the necessity to discover additional therapy modalities that could prolong survival and relieve symptoms.

In H3-K27M-mutant DMGs, as well as in many other malignancies, high MYC activity is correlated with poor prognosis and therefore considered as an attractive therapeutic target21,83. Nevertheless, direct targeting of MYC is highly challenging, functional and structural wise. MYC signaling, while amplified in cancer cells, is also essential to normal cell function84. Moreover, as a transcription factor, MYC lacks traditionally druggable binding pockets, hindering the development of direct small molecule binders25. To overcome this, in this study we exploited Sulfopin, a novel, potent and selective PIN1 inhibitor26. Pre-clinical studies have shown in-vivo activity and negligible toxicity of Sulfopin in murine models of PDAC and neuroblastoma26. While PIN1 is not overexpressed in DMG tumors, we show that Sulfopin, although not inhibiting MYC directly, downregulates MYC target genes in patient-derived DMG cells and affects cell survival in a H3-K27M-dependant manner. These results are in line with the contribution of MYC signaling to H3-K27M-driven tumorigenesis23,24, underscoring the therapeutic potential of Sulfopin for these aggressive brain tumors. Notably, despite a significant reduction in tumor size in-vivo, the combined treatment did not increase mice survival. This is perhaps due to the relatively large tumors already formed at the onset of treatment, pointing to the need for further optimization of the therapeutic regime.

We showed that combining Sulfopin treatment with the HDAC inhibitor Vorinostat reduced cell viability by 80% in several tumor-derived DMG lines originating from different patients. Importantly, the combination resulted in an additive effect not only on cell viability, but also on transcriptional programs. A key oncogenic pathway that was robustly downregulated in the combined treatment, in-vitro and in-vivo, is the mTOR signaling pathway. Preclinical and clinical data support a role for mTOR in gliomagenesis85. Specifically, AKT gain or PTEN loss is detected in approximately 70% of DMG tumors, and a recent study showed high expression of mTOR in DMGs. Thus, the PTEN/AKT/mTOR pathway is suggested to play a key role in this disease86–89. Attempts to restrict mTOR signaling using first generation mTOR inhibitors primarily inhibited mTORC1, which often caused upregulation of mTORC290. The combined treatment with Sulfopin and Vorinostat led to robust inhibition of several target proteins in the mTOR signaling pathway, likely via alternative pathways. One potential candidate is the mTOR-dependent cell-growth through MYC regulation, which may be impaired by Sulfopin treatment, inhibiting MYC transcriptional activity91–94. Of note, drug agents targeting both mTORC1 and mTORC2 were shown to be highly effective in DMG cells as well as in murine models95,96.

Expression of the H3-K27M oncohistone is associated with a prominent increase in H3K27ac levels11–13,16. Among the loci that accumulate this modification are repetitive elements and endogenous retroviruses (ERVs), resulting in their aberrant expression, potentially inducing innate immune responses12. However, H3-K27M cells do not show an increase in interferon stimulated genes (ISGs), preassembly due to high levels of MYC, which is potent antagonist of interferon responses12,97. Thus, dual targeting of both HDAC and MYC pathways has the potential of inducing robust expression of ISGs, thus promoting tumor cell death mediated by viral-mimicry98.

The high levels of H3K27ac in H3-K27M cells was also linked to activation of super-enhancers, which regulate the expression of oncogenes and genes that are associated with an undifferentiated state79. Here, we show that H3-K27M cells are vulnerable to transcriptional disruption mediated by the combined treatment with Sulfopin and Vorinostat. While treating cells with Vorinostat alone increased H3K27ac levels, as expected, the combined treatment attenuated the accumulation of this mark on promoters and enhancers of prominent oncogenes, resulting in their decreased expression. This Sulfopin-mediated attenuation could potentially result from downregulation of MYC targets, as it was recently shown that MYC overexpression results in H3K27ac gain on super-enhancers99.

In summary, we propose a novel therapeutic strategy targeting epigenetic and transcriptional pathways in H3-K27M-mutant DMG tumors. HDAC inhibition, in combination with the novel MYC inhibitor, inhibited oncogenic transcriptional signatures and reduced tumor growth in-vivo. Our findings emphasize epigenetic dependencies of H3-K27M driven gliomas and highlight the therapeutic potential of combined treatments for this aggressive malignancy.

Supporting information

eLife Assessment

This is valuable work showing that a combination of drugs can reduce growth of Diffuse midline gliomas (clinically classified as DMG, H3 K27M-mutant) when applied in vitro and in tumor xenografts in mice. It is a significant first step towards understanding how these drugs work, and provides convincing results to encourage future pre-clinical studies. Further rationale on how doses for specific drugs were chosen, directly demonstrating a survival benefit, or implicating the Pin1 pathway components mechanistically, would make the manuscript stronger.

Reviewer #1 (Public Review):

Summary:

This is an interesting study that utilizes a novel epigenome profiling technology (single molecule imaging) in order to demonstrate its utility as a readout of therapeutic response in multiple DIPG cell lines. Two different drugs were evaluated, singly and in combination. Sulfopin, an inhibitor of a component upstream of the MYC pathway, and Vorinostat, an HDAC inhibitor. Both drugs sensitised DIPG cells, but high (>10 micromolar) concentrations were needed to achieve half-maximal effects. The combination seemed to have some efficacy in vivo, but also produced debilitating side-effects that precluded the measurement of any survival benefit.

Strengths:

Interesting use of a novel epigenome profiling technology (single molecule imaging).

Weaknesses:

The use of this novel imaging technology ultimately makes up only a minor part of the study. The rest of the results, i.e. DIPG sensitivity to HDAC and MYC pathway inhibition, have already been demonstrated by others (Grasso Monje 2015; Pajovic Hawkins 2020, among others). The drugs have some interesting opposing effects at the level of the epigenome, demonstrated through CUT&RUN, but this is not unexpected in any way. The drugs evaluated here also didn't have higher efficacy, or efficacy at especially low concentrations, than inhibitors used in previous reports. The combination therapy attempted here also caused severe side effects in mice (dehydration/deterioration), such that an effect on survival could not be determined. I'm not sure this study advances knowledge of targeted therapy approaches in DIPGs, or if it iterates on previous findings to deliver new, or more efficient, mechanistic or therapeutic/pharmaclogic insights. It is a translational report evaluating two drugs singly and in combination, finding that although they sensitise cells in vitro, efficacy in vivo is limited at best, as this particular combination cannot progress to human translation.

Reviewer #2 (Public Review):

Summary:

The study by Algranati et al. introduces an exciting and promising therapeutic approach for the treatment of H3-K27M pediatric gliomas, a particularly aggressive brain cancer predominantly affecting children. By exploring the dual targeting of histone deacetylases (HDACs) and MYC activation, the research presents a novel strategy that significantly reduces cell viability and tumor growth in patient-derived glioma cells and xenograft mouse models. This approach, supported by transcriptomic and epigenomic profiling, unveils the potential of combining Sulfopin and Vorinostat to downregulate oncogenic pathways, including the mTOR signaling pathway. While the study offers valuable insights, it would benefit from additional clarification on several points, such as the rationale behind the dosing decisions for the compounds tested, the specific contributions of MYC amplification and H3K27me3 alterations to the observed therapeutic effects, and the details of the treatment protocols employed in both in-vitro and in-vivo experiments.

Clarification is needed on how doses were selected for the compounds in Figure S2A and throughout the study. Understanding the basis for these choices is crucial for interpreting the results and their potential clinical relevance. IC50s are calculated for specific patient derived lines, but it is not clear how these are used for selecting the dose.

The introduction mentions MYC amplification in high-grade gliomas. It would be beneficial if the authors could delineate whether the models used exhibit varying degrees of MYC amplification and how this factor, alongside differences in H3K27me3, contributes to the observed effects of the treatment.

In Figure 2A, the authors outline an optimal treatment timing for their in vitro models, which appears to be used throughout the figure. It would be helpful to know how this treatment timing was selected and also why Sulfopin is dosed first (and twice) before the vorinostat. Was this optimized?

It should be clarified whether the dosing timeline for the combination drug experiments in Figure 3 aligns with that of Figure 2. This information is also important for interpreting the epigenetic and transcriptional profiling and the timing should be discussed if they are administered sequentially (also shown in Figure 2A).I have the same question for the mouse experiments in Figure 4.

The authors mention that the mice all had severe dehydration and deterioration after 18 days. It would be helpful to know if there were differences in the side effects for different treatment groups? I would expect the combination to be the most severe. This is important in considering the combination treatment.

Minor Points:

(1) For Figure 1F, reorganizing the bars to directly compare the K27M and KO cell lines at each dose would improve readability of this figure.

(2) In Figure 4D, it would be helpful to know how many cells were included (or a minimum included) to calculate the percentages.

Reviewer #3 (Public Review):

Summary:

The authors use in vitro grown cells and mouse xenografts to show that a combination of drugs, Sulfopin and Vorinostat, can impact the growth of cells derived from Diffuse midline gliomas, in particular the ones carrying the H3 K27M-mutations (clinically classified as DMG, H3 K27M-mutant). The authors use gene expression studies, and chromatin profiling to attempt to better understand how these drugs exert an effect on genome regulation. Their main findings are that the drugs reduce cell growth in vitro and in mouse xenografts of patient tumours, that DMG, H3 K27M-mutant tumours are particularly sensitive, identify potential markers of gene expression underlying this sensitivity, and broadly characterize the correlations between chromatin modification changes and gene expression upon treatment, identifying putative pathways that may be affected and underlie the sensitive (and thus how the drugs may affect the tumour cell biology).

Strengths:

It is a neat, mostly to-the-point work without exploring too many options and possibilities. The authors do a good job not overinterpreting data and speculating too much about the mechanisms, which is a very good thing since the causes and consequences of perturbing such broad epigenetic landscapes of chromatin may be very hard to disentangle. Instead, the authors go straight after testing the performance of the drugs, identifying potential markers and characterizing consequences.

Weaknesses:

If anything, the experiments done on Figure 3 could benefit from an additional replicate.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This is an interesting study that utilizes a novel epigenome profiling technology (single molecule imaging) in order to demonstrate its utility as a readout of therapeutic response in multiple DIPG cell lines. Two different drugs were evaluated, singly and in combination. Sulfopin, an inhibitor of a component upstream of the MYC pathway, and Vorinostat, an HDAC inhibitor. Both drugs sensitized DIPG cells, but high (>10 micromolar) concentrations were needed to achieve half-maximal effects. The combination seemed to have some efficacy in vivo, but also produced debilitating side-effects that precluded the measurement of any survival benefit.

We thank the reviewer for deeply evaluating our work and acknowledging the use of multiple experimental strategies to explore the effect of combination therapy on DMG cells. Of note, all mice in our experiment experienced deterioration (including the control mice and those treated with single agents). Thus, it is not the combination of drugs that led to the debilitating side-effects; the mice deteriorated due to the extremely aggressive tumor cells, forming relatively large tumors prior to the treatment onset, calling for further optimization of the therapeutic regime.

We modified the text in the results section to clarify this point (lines 238-241): “This rapid deterioration is likely a result of the aggressiveness of the transplanted tumors and does not represent side effects of the treatment, as mice from all groups, including the non-treated mice, showed similar signs of deterioration”.

We also elaborate on this in the discussion (lines 272-276): “Notably, despite a significant reduction in tumor size in-vivo, the combined treatment did not increase mice survival. This is perhaps due to the relatively large tumors already formed at the onset of treatment, leading to rapid deterioration of mice in all experimental groups. Thus, further optimization of the modeling system and therapeutic regime is needed.” We truly hope that further studies will allow better assessment of this drug combination in various models.

Strengths:

Interesting use of a novel epigenome profiling technology (single molecule imaging).

Weaknesses:

The use of this novel imaging technology ultimately makes up only a minor part of the study. The rest of the results, i.e. DIPG sensitivity to HDAC and MYC pathway inhibition, have already been demonstrated by others (Grasso Monje 2015; Pajovic Hawkins 2020, among others). The drugs have some interesting opposing effects at the level of the epigenome, demonstrated through CUT&RUN, but this is not unexpected in any way. The drugs evaluated here also didn't have higher efficacy, or efficacy at especially low concentrations, than inhibitors used in previous reports. The combination therapy attempted here also caused severe side effects in mice (dehydration/deterioration), such that an effect on survival could not be determined. I'm not sure this study advances knowledge of targeted therapy approaches in DIPGs, or if it iterates on previous findings to deliver new, or more efficient, mechanistic or therapeutic/pharmaclogic insights. It is a translational report evaluating two drugs singly and in combination, finding that although they sensitise cells in vitro, efficacy in vivo is limited at best, as this particular combination cannot progress to human translation.

We thank the reviewer for pointing out the strengths and weaknesses of our work. As far as we know, while many studies demonstrated upregulation of the MYC pathway in DIPG, this is the first study that shows inhibition of this pathway (via PIN1) as a therapeutic strategy. While it is clear from the literature that MYC inhibition may pose therapeutic benefit, the development of potent MYC inhibitors is highly challenging due to its structure and cellular localization. Of note, in the 2020 paper, Pajovic and colleagues inhibited MYC by transfecting the cells with a plasmid expressing a specific inhibitory MYC peptide (Omomyc); while this strategy works well for cell cultures, the clinical translation requires different delivery strategies. Sulfopin is a small molecule inhibitor that can be used in-vivo and potentially in clinical studies. Thus, we believe that our study offers a novel strategy, as well as mechanistic insights, regarding the potential use of Sulfopin and Vorinostat to treat DIPG.

As noted above, the combination therapy did not cause side effects, but rather the aggressiveness of the tumors. We did not notice specific toxicity in the mice treated with Sulfopin alone, or the combined treatment. Furthermore, Dubiella et al. extensively examined toxicity issues and did not observe adverse effects or weight loss when administrating Sulfopin in a dose of 40 mg kg–1.

Optimization of the model and treatment regime (# of cells injected, treatment starting point, etc.) may have allowed us to reveal survival benefits. Yet, these are highly complicated and expensive experiments; unfortunately, we did not have the resources to perform them within the scope of this revision. Importantly, within the current manuscript, we show the effect of this drug combination in reducing the growth of DMG cells in-vitro and in-vivo, laying the framework for follow-up exploration in future studies. Furthermore, the epigenetic and transcriptomic profiling shed light on the molecular mechanisms that drive these aggressive tumors.

Reviewer #2 (Public Review):

Summary:

The study by Algranati et al. introduces an exciting and promising therapeutic approach for the treatment of H3-K27M pediatric gliomas, a particularly aggressive brain cancer predominantly affecting children. By exploring the dual targeting of histone deacetylases (HDACs) and MYC activation, the research presents a novel strategy that significantly reduces cell viability and tumor growth in patient-derived glioma cells and xenograft mouse models. This approach, supported by transcriptomic and epigenomic profiling, unveils the potential of combining Sulfopin and Vorinostat to downregulate oncogenic pathways, including the mTOR signaling pathway. While the study offers valuable insights, it would benefit from additional clarification on several points, such as the rationale behind the dosing decisions for the compounds tested, the specific contributions of MYC amplification and H3K27me3 alterations to the observed therapeutic effects, and the details of the treatment protocols employed in both in-vitro and in-vivo experiments.

We thank the reviewer for evaluating our work and recognizing its potential for the DMG research field. We address in detail below the important comments regarding the treatment protocols and dosing decisions.

Clarification is needed on how doses were selected for the compounds in Figure S2A and throughout the study. Understanding the basis for these choices is crucial for interpreting the results and their potential clinical relevance. IC50s are calculated for specific patient derived lines, but it is not clear how these are used for selecting the dose.

We thank the reviewer for these important comments. For the epigenetic drugs shown in Figure S2A, we followed published experimental setups; for EPZ6438, GSKJ4, Vorinostat and MM-102 we chose the treating concentrations according to Mohammad et al. 2017, Grasso et al. 2015 and Furth et al. 2022, accordingly. For Sulfopin, we conducted a dedicated dose curve analysis (shown in Figure 1E), indicating only a mild effect on viability and relatively high IC-50 values as a single agent. Since we aimed to test the ability of a combined treatment to additively reduce viability, we used a sub-IC50 concentration for Sulfopin in these experiments. We added this information in lines 123 and 131-132.

Finally, following the results obtained in the experiment shown in Figure S2A, we conducted a full dose-curve analysis of the combined treatment in multiple DMG patient-derived cells (figure 2B and S2C), to identify a combination of concentrations that provides an additive effect (as indicated by BLISS index in figure 2C and S2E). Of note, for downstream analysis of the molecular mechanisms underlying the treatment response (RNAseq and Cut&Run), we intentionally used concentrations that provide an additive BLISS index, but do not completely abolish the culture, to allow for cellular analysis (i.e. 10uM Sulfopin and 1uM Vorinostat).

The introduction mentions MYC amplification in high-grade gliomas. It would be beneficial if the authors could delineate whether the models used exhibit varying degrees of MYC amplification and how this factor, alongside differences in H3K27me3, contributes to the observed effects of the treatment.

The reviewer highlights an important part of our study relating to the MYC-dependent sensitivity of the proposed treatment combination. Since high expression of MYC can be mediated by different molecular mechanisms and not only genomic amplification, we directly quantified mRNA levels of MYC by qPCR (shown in figure S2G) in order to explore its relationship with cellular response to Sulfopin and Vorinostat. Indeed, cultures that express high levels of MYC mRNA were more sensitive to Sulfopin treatment alone (figure S1P) and to the combined treatment (figure 2D-E). We also relate to these findings in lines 103-106 and 142-147 of the results section. Importantly, in cultures that express high levels of MYC (SU-DIPG13 as an example), we see downregulation of MYC targets upon the combined treatment, supporting the notion that this treatment affects viability by attenuation of MYC signaling.

In Figure 2A, the authors outline an optimal treatment timing for their in vitro models, which appears to be used throughout the figure. It would be helpful to know how this treatment timing was selected and also why Sulfopin is dosed first (and twice) before the vorinostat. Was this optimized?

As PIN1 regulates the G2/M transition, its inhibition by Sulfopin delays cell cycle progression (Yeh et al. 2007). Thus, in order to observe a strong viability difference in culture, a prolonged treatment period of 8-9 days is required (Dubiella et al., 2021). To maintain an active concentration of the drug during this long time period, we added a Sulfopin pulse (2nd dose) to achieve a stronger effect on cell viability. We and others noticed that, unlike Sulfopin, the effect of Vorinostat on viability is rapid and can be clearly seen after 2-3 days of treatment. Thus, we added this drug only after the 2nd dose of Sulfopin. We now relate to the mode of action of Sulfopin in lines 79-81.

It should be clarified whether the dosing timeline for the combination drug experiments in Figure 3 aligns with that of Figure 2. This information is also important for interpreting the epigenetic and transcriptional profiling and the timing should be discussed if they are administered sequentially (also shown in Figure 2A).I have the same question for the mouse experiments in Figure 4.

The reviewer is correct that this information is critical for evaluating the results. In order to link the expression changes to the epigenetic changes, we kept the same experimental conditions in both the Cut&Run and RNA-seq experiments (shown in figures 2-3). We added this information to the text in line 184.

For the in-vivo studies of HDAC inhibition (Figure 4), we followed published protocols (Ehteda et al. 2021). In these experiments both drugs were administrated simultaneously every day. We added this information to the text in line 231-232. It may be that changing the admission regime may improve the efficacy of the drug combination, which remains to be tested in future studies.

The authors mention that the mice all had severe dehydration and deterioration after 18 days. It would be helpful to know if there were differences in the side effects for different treatment groups? I would expect the combination to be the most severe. This is important in considering the combination treatment.

As noted in our response to Reviewer #1, all mice in our experiment experienced deterioration (including the control mice and those treated with single agents- we could not observe any differences between the groups). This is due to the extremely aggressive tumor cells, forming relatively large tumors prior to the treatment onset, calling for further optimization of the system and therapeutic regime (# of cell injected, treatment starting point, etc.). Unfortunately, this model is very challenging (especially the injection of cells to the pons of the mice brains, which requires unique expertise and is associated with mortality of some of the mice). Thus, these are highly complicated and expensive experiments; unfortunately, we did not have the resources to repeat and optimize the treatment protocol within the scope of this revision. Of note, Dubiella et al. extensively examined toxicity issues and did not observe adverse effects or weight loss when administrating Sulfopin in a dose of 40 mg kg–1. In our model, the side effects were caused by the tumors rather than the drugs.

Minor Points:

(1) For Figure 1F, reorganizing the bars to directly compare the K27M and KO cell lines at each dose would improve readability of this figure.

We have changed figure 1F as the reviewer suggested.

(2) In Figure 4D, it would be helpful to know how many cells were included (or a minimum included) to calculate the percentages.

We added the number of H3-K27M positive cells detected per FOV to the figure legend and method section (n=13-198 cells per FOV). Of note, while we analyzed similar-sized FOVs, the number of tumor cells varied between the groups, with the treated group presenting a lower number of H3-K27M cells (due to the effect of the treatment on tumor growth). To account for this difference, we calculated the portion of mTOR-positive cells out of the tumor cells.

Reviewer #3 (Public Review):

Summary:

The authors use in vitro grown cells and mouse xenografts to show that a combination of drugs, Sulfopin and Vorinostat, can impact the growth of cells derived from Diffuse midline gliomas, in particular the ones carrying the H3 K27M-mutations (clinically classified as DMG, H3 K27M-mutant). The authors use gene expression studies, and chromatin profiling to attempt to better understand how these drugs exert an effect on genome regulation. Their main findings are that the drugs reduce cell growth in vitro and in mouse xenografts of patient tumours, that DMG, H3 K27M-mutant tumours are particularly sensitive, identify potential markers of gene expression underlying this sensitivity, and broadly characterize the correlations between chromatin modification changes and gene expression upon treatment, identifying putative pathways that may be affected and underlie the sensitive (and thus how the drugs may affect the tumour cell biology).

Strengths:

It is a neat, mostly to-the-point work without exploring too many options and possibilities. The authors do a good job not overinterpreting data and speculating too much about the mechanisms, which is a very good thing since the causes and consequences of perturbing such broad epigenetic landscapes of chromatin may be very hard to disentangle. Instead, the authors go straight after testing the performance of the drugs, identifying potential markers and characterizing consequences.

Weaknesses:

If anything, the experiments done on Figure 3 could benefit from an additional replicate.

We thank the reviewer for evaluating our work, and for the positive and insightful comments.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Perhaps a more substantial drug screen, or CRISPR screen, that utilises single molecule imaging as a readout would identify pharmacologic candidates that are either more effective, or novel.

While out of scope for the current study, this is a highly interesting suggestion, which will be considered in future studies. Here, we focused on the potential use of the novel MYC inhibitor, Sulfopin. While the dependency of DMG cells on MYC signaling has been documented, to the best of our knowledge, pharmacological inhibition of MYC has not been tested for this disease due to the severe lack of potent MYC inhibitors. We show preliminary evidence for the use of this inhibitor, in combination with HDAC inhibition, to attenuate DMG growth in-vitro and in-vivo.

Reviewer #2 (Recommendations For The Authors):

In Figure 1B, it is hard to tell if there are error bars for HSP90 and E2F2. Is there a potential error here? Seems unlikely to not have an error with a RT-qPCR?

We thank the reviewer for the careful evaluation of the figures. We included error bars for all genes shown in Figure 1B. We have now increased the line width with the hope of making this information more accessible. As stated in the figure legend, these error bars represent the standard deviation of two technical repeats.

I noticed that many experiments only had technical replicates. Incorporating biological (independent) replicates, where feasible, would strengthen the study's findings.

We agree with the reviewer regarding the importance of biological replicates. While some of the panels present error estimates based on technical repeats, the main results were repeated independently with complementary approaches or various biological systems for validation.

The RNAseq analysis presented in figure 1 was conducted in triplicates and then independently validated by qPCR (Figure 1A-B). Similarly, the transcriptomic analysis presented in figures 2G-I was verified by both western blot (figure 2J) and qPCR (figure S2O). Of note, this later validation was conducted for two different DMG-patient derived cultures.

To verify the robust effects on cellular viability, we analyzed the response to each drug and the combination on eight different DMG-patient-derived cultures, each representing a completely independent experiment. We show very similar trends in response to treatment between cultures that share the same H3-K27M variant. Thus, while for each culture technical repeats are shown, we provide multiple, independent repeats by examining the different cultures. Similarly, in figure 1F we examined the dependency of Sulfopin treatment on the expression of the H3-K27M oncohistone in two independent isogenic systems.

Reviewer #3 (Recommendations For The Authors):

A few questions and suggestions:

(1) To avoid confusion is important to state if the cells used in each experiment are or not K27M mutants (e.g. SU-DIPG13 on line 63).

We thank the reviewer for pointing this out and have now added this information when appropriate across the manuscript.

1. Line 72 - confirming epigenetic silencing of these genes upon PIN1 inhibition (Fig. 1C, S1D)

Considering that the mechanism of down regulation of MYC targets is likely H3K27me3-independent if it is also happening in DMG H3 K27M-mutants (high H3K27me3 here may rather be a consequence of less MYC binding?), I would strike this sentence out and just point out the correlation between lower expression and higher H3K27me3.

We agree with the reviewer that the exact molecular mechanism mediating the silencing is yet to be characterized. We have modified the text in line 72 accordingly.

1. (line 78) Are MYC targets also down regulated in Sulfopin treated DMG, H3 K27M-mutant lines? Any qPCR or previously done RNA-seq data to use?

In addition to the extensive analysis done on SU-DIPG13 cells (Figure 1 and S1), in light of the reviewer`s comment we examined specific MYC targets in an additional H3-K27M mutant DMG culture (SU-DIPG6) treated with Sulfopin, followed by qPCR. We observed a mild reduction in two prominent targets, E2F2 and mTOR (new figure S1D). Unfortunately, within this study, we only conducted full RNA-sequencing analysis on SU-DIPG13 cells treated with Sulfopin, and thus, we could not examine the global effect of Sulfopin on the transcriptome of other DMG cultures. This will, of course, be of high interest for future studies.

Version 2

Abstract

Abstract

Diffuse midline gliomas (DMG) are aggressive and fatal pediatric tumors of the central nervous system that are highly resistant to treatments. Lysine to methionine substitution of residue 27 on histone H3 (H3-K27M) is a driver mutation in DMGs, reshaping the epigenetic landscape of these cells to promote tumorigenesis. H3-K27M gliomas are characterized by deregulation of histone acetylation and methylation pathways, as well as the oncogenic MYC pathway. In search of effective treatment, we examined the therapeutic potential of dual targeting of histone deacetylases (HDACs) and MYC in these tumors. Treatment of H3-K27M patient-derived cells with Sulfopin, an inhibitor shown to block MYC-driven tumors in-vivo, in combination with the HDAC inhibitor Vorinostat, resulted in substantial decrease in cell viability. Moreover, transcriptome and epigenome profiling revealed synergistic effect of this drug combination in downregulation of prominent oncogenic pathways such as mTOR. Finally, in-vivo studies of patient-derived orthotopic xenograft models showed significant tumor growth reduction in mice treated with the drug combination. These results highlight the combined treatment with PIN1 and HDAC inhibitors as a promising therapeutic approach for these aggressive tumors.

Introduction

Diffuse midline gliomas (DMGs) are fatal tumors of the central nervous system that occur primarily in children1,2. Years of clinical trials have revealed that conventional chemotherapy is ineffective, and with no option for complete resection, DMG is now the leading cause of brain tumor-related death in children3,4. Sequencing studies showed that up to 80% of the diagnosed children exhibit a lysine 27 to methionine (H3-K27M) mutation in one of the genes encoding histone H3.3 and H3.1 (clinically classified as DMG, H3 K27M–mutant). Numerous subsequent studies revealed pronounced alterations in the epigenetic landscape driven by this histone mutation5–8. Among these global epigenetic alterations are a drastic loss of H3 lysine 27 tri-methylation (H3K27me3), concomitant with an increase in H3 lysine 27 acetylation (H3K27ac). These epigenetic alterations were shown to affect transcriptional programs and support tumorigenesis 9–14.

Detailed characterization of the modes of chromatin deregulation in these tumors pointed toward targeting epigenetic pathways as a promising therapeutic approach15. Specifically, pervasive H3K27 acetylation patterns observed in H3-K27M DMGs, along with enrichment of H3K27ac specifically on mutant nucleosomes11–13,16, underscored the histone acetylation pathway as a potential drug target for this cancer. Indeed, pan-histone deacetylase inhibitors (HDACi), such as Vorinostat and Panobinostat, were found to inhibit growth and to restore gene expression alterations observed in H3-K27M malignant gliomas17–19. However, recent studies demonstrated that H3-K27M cells develop resistance to HDACi treatment17,19, stressing the importance of identifying novel drug combinations to induce synergistic inhibition of oncogenic pathways.

The MYC oncogenic pathway is frequently deregulated in diverse cancers, and thus stands as a prominent therapeutic target20. High-level gene amplification of MYC is recurrently observed in pediatric high-grade gliomas21,22. H3-K27M gliomas show high expression of MYC and MYC target genes, due to both epigenetic alterations and structural variants, resulting in a viability dependency on MYC signaling12,23,24. Unfortunately, directly targeting MYC in tumors is extremely challanging25. Dubiella et al. have recently developed a unique inhibitor targeting Peptidyl-prolyl isomerase NIMA-interacting-1 (PIN1), an upstream regulator of MYC activation26. PIN1 itself is frequently upregulated in many types of cancers, resulting in sustained proliferative signaling and tumor growth27,28. The PIN1 inhibitor, Sulfopin, was shown to downregulate MYC target genes in-vitro and to block MYC-driven tumors in murine models of neuroblastoma26.

To address the increasing need for an effective combination therapy for H3-K27M DMGs, we tested whether dual targeting of histone acetylation and MYC activation, using Vorinostat and Sulfopin, would yield beneficial outcome. We show that the combined treatment substantially reduces viability of patient-derived glioma cells, in concordance with inhibition of prominent oncogenic pathways, including the mTOR pathway. Treatment of H3-K27M DMG xenograft mice models with this drug combination led to reduced tumor growth and confirmed downregulation of mTOR in-vivo. Together, these results suggest the combination of PIN1 and HDAC inhibitors as a promising therapeutic strategy for H3-K27M gliomas.

Results

Sulfopin inhibits MYC targets and reduces viability of H3-K27M glioma cells

In light of the important role of MYC in H3-K27M DMGs, we set to explore whether the PIN1 inhibitor, Sulfopin, may affect MYC signaling in these cells and thus confer therapeutic potential. Sulfopin dependent inhibition of PIN1 was shown to downregulate the expression of MYC target genes in several cancer cell lines26. Indeed, RNA sequencing analysis of H3-K27M mutant patient-derived glioma cells (SU-DIPG13) treated with Sulfopin revealed strong enrichment of MYC targets among the downregulated genes (Fig. 1A). The genes downregulated by Sulfopin were also enriched for MYC-bound genes identified in several non-glioma cell lines, supporting a role for PIN1 in activation of the MYC pathway also in these glioma cells (Fig. S1A-B). Of note, Sulfopin was developed as a covalent inhibitor of PIN1, and did not lead to PIN1 degradation, as previously shown (Fig. S1C) 26. We further validated the decreased expression of prominent MYC targets27,29,30 upon Sulfopin treatment by RT-qPCR analysis (Fig. 1B, S1D). Interestingly, Cut&Run analysis31 revealed an increase in H3K27me3 levels, indicative of repressed chromatin32, around the transcriptional start sites (TSS) of MYC target genes, correlating with silencing of these genes upon PIN1 inhibition (Fig. 1C, S1E). Of note, global H3K27me3 distribution on all TSS and gene-body regions was not affected by Sulfopin treatment, supporting a specific increase in H3K27me3 on the TSS of MYC target genes (Fig. S1F-G). Furthermore, Sulfopin-unique H3K27me3 peaks (i.e., peaks that were found only in the Sulfopin-treated cells, and not in DMSO-treated cells) were enriched for MYC target genes (Fig. 1D).

Sulfopin inhibits MYC signaling and reduces viability of DMG cells in a H3-K27M-dependent manner

(A) Functional enrichment analysis on significantly downregulated genes in SU-DIPG13 cells treated with 10uM Sulfopin for 12 hours, compared to DMSO. Enrichr algorithm123 was used to compare downregulated genes against the Molecular Signatures Database (MSigDB) hallmark geneset119. Dashed line denotes adjusted p-value = 0.05. MYC targets are significantly enriched among the Sulfopin downregulated genes.

(B) RT-qPCR analysis of selected MYC target genes in SU-DIPG13 cells treated with 10uM Sulfopin for 12 hours, compared to DMSO. Fold change between Sulfopin and DMSO treated cells was calculated and the mean ± SD of two technical repeats is shown.

(C) H3K27me3 Cut&Run read coverage over MYC target genes (‘MYC Targets V1’ hallmark geneset119, n=200), in SU-DIPG13 cells treated with 10uM Sulfopin for 8 days compared to DMSO. Sulfopin treatment increases H3K27me3 levels on the TSS of MYC target genes.

(D) Functional enrichment analysis of the genes associated with Sulfopin-unique H3K27me3 peaks, in SU-DIPG13 cells treated as in C. Dashed line denotes adjusted p-value = 0.05. MYC target geneset (‘MYC Targets V1’ hallmark geneset119) is mildly enriched among these genes, with adjusted p-value of 0.077.

(E) Cell viability, as measured by CellTiterGlo, of eight DMG cultures (H3.3K27M: SU-DIPG13, SU-DIPG6, SU-DIPG17, SU-DIPG25 and SU-DIPG50. H3.1K27M: SU-DIPG36, SU-DIPG38 and SU-DIPG21), treated with Sulfopin for eight days with pulse at day four, compared to DMSO. Mean±SD of two technical replicates is shown. Logarithmic scale is used for the x-axis. Sulfopin treatment led to a mild reduction in cell viability in all H3-K27M glioma cultures.

(F) Cell viability, as measured by CellTiterGlo, of two isogenic DMG cell lines (SU-DIPG13 and BT245) in which the mutant histone was knocked-out (KO), treated with the indicated concentration of Sulfopin for 8 days, compared to DMSO. For each cell line and concentration, the fold change in viability between Sulfopin and DMSO treated cells is shown. For SU-DIPG13-mean ± SE of at least two independent experiments is shown. For BT245-mean± SD of three technical replicates is shown. H3-K27M glioma cells show higher sensitivity to Sulfopin treatment compared to the KO cells. *P < 0.05; **P < 0.01 (two-sample t-test over all technical replicates). Significance adjusted after Bonferroni correction.

We next explored the effect of Sulfopin on the viability of eight patient derived DMG cultures, harboring either the H3.3-K27M or H3.1-K27M mutation. Inhibition of PIN1 by Sulfopin was previously shown to delay cell cycle progression and thus reduce cell viability after extended treatments of 6-8 days 26. In line with downregulation of MYC signaling, Sulfopin treatment led to a mild reduction in cell viability, conserved across the different cultures (Fig. 1E). To examine whether H3-K27M mutation renders DMG cells more sensitive to Sulfopin treatment, we took advantage of two sets of isogenic H3-K27M mutant patient-derived DMG cell-lines (SU-DIPG13, BT245) in which the H3F3A-K27M mutant allele has been knocked-out (KO). In both lines, the reduction in cell viability was partially rescued in the KO cells (Fig. 1F), suggesting higher sensitivity of H3-K27M mutant cells to MYC inhibition by Sulfopin. These results are in line with previous reports showing H3-K27M-dependent activation of the MYC/RAS axis in mouse and human cells12,23,24. Of note, DMG cells showed greater sensitivity to the inhibitor compared to human astrocytic-like cells (Fig. S1H).

To further explore the status of MYC and PIN1 as indicators of response to Sulfopin inhibition, we analyzed published transcriptomic dataset of DMG tumors35. Interestingly, while DMG samples showed significantly higher expression of MYC and its target genes compared to normal samples, PIN1 expression was lower in the cancer compared to matched normal tissue (Fig. S1I-K). Furthermore, H3-K27M mutant tumors showed higher expression of MYC, compared to H3 WT tumors, while PIN1 levels were reduced across samples22 (Fig. S1L-M). These results suggest that MYC expression levels, and not PIN1, may determine the sensitivity of H3-K27M cells to Sulfopin treatment. To examine this notion, we measured the expression levels of MYC, as well as the doubling time of the different DMG cultures. The response to Sulfopin treatment (as reflected by IC50 measurements) correlated with the cells’ doubling time (Fig. S1N). Cells that divide faster were more sensitive to Sulfopin treatment, in line with PIN1 role in regulating cell cycle progression, and specifically G2/M transition36,37. As expected, the doubling time negatively correlated with MYC expression, suggesting that MYC levels may predict the response to Sulfopin (Fig S1O). Indeed, MYC levels negatively correlated with the IC50 of Sulfopin; thus, cells relaying on high MYC activity showed higher sensitivity to Sulfopin (Fig S1P).

H3K27me3 levels were shown to be drastically deregulated in DMG cells expressing H3-K27M, with global loss of this modification across the genome, and retention of residual H3K27me3 at CpG islands (CGIs). These residual H3K27me3 levels were shown to be necessary to silence tumor suppressor genes and maintain cell proliferation in DMG cells33,34. Interestingly, Sulfopin treatment resulted in reduction of H3K27me3 signal over CGIs, perhaps contributing to Sulfopin-dependent growth inhibition in these cells (Fig. S1Q). Taken together, our results indicate that PIN1 inhibition by Sulfopin downregulated MYC target genes, reduced H3K27me3 levels at CGIs, and led to reduce viability of DMG cells in a H3-K27M-dependant manner.

Combination of Sulfopin and Vorinostat elicits robust inhibition of oncogenic pathways and an additive effect on cell viability

Inspired by the positive outcome of combinatorial drug therapy in H3-K27M DMG models17,38, we aimed to explore whether combining Sulfopin treatment with an epigenetic drug will further reduce cell viability. H3-K27M gliomas are characterized by deregulation of epigenetic pathways affecting post-translational modifications of lysine 27 (methylation and acetylation) and lysine 4 (methylation) on histone H39–12,16,39. Thus, we explored the effects of combining sub-IC50 concentration of Sulfopin with four epigenetic inhibitors targeting critical components associated with these modifications: (1) EPZ-6438, an inhibitor of EZH2, the catalytic subunit of the PRC2 complex that deposits H3K27me340. H3-K27M cells were reported to be sensitive to EZH2 inhibition as they rely on the residual H3K27me3 levels to suppress key tumor-suppressor genes13,33,41,42; (2) GSK-J4, an inhibitor of the H3K27me3 demethylase JMJD343, shown to rescue H3K27me3 levels in H3-K27M gliomas17,44; (3) MM-102, an inhibitor of the H3K4me3 methyltransferase MLL140, shown to reduce viability of H3-K27M-mutant cells16; and (4) Vorinostat, an HDAC inhibitor that was shown to have clinical benefit in H3-K27M gliomas17–19. Treatment dosages of these drugs were set according to previous studies 33,17,16. We found that the combination of Sulfopin and Vorinostat (HDACi) had the strongest effect, reducing viability by 80% (Fig. S2A).

Next, we adjusted treatment protocols to account for the optimal treatment period for each drug, and examined the effect of this drug combination on the viability of eight H3-K27M mutant patient-derived DMG cultures, across different dosages (Fig. 2A-B and S2B-D). For each pair of concentrations, we calculated a BLISS index; an additive effect of the two drugs was determined as BLISS index ∼145,46 (Fig. 2C, S2E). While the combined treatment was effective in all the cultures, cells harboring H3.3-K27M showed higher sensitivity (Fig. 2D, S2F). Moreover, while an additive effect of Sulfopin and Vorinostat was detected in all the H3.3-K27M DMG cultures in most of the drug doses, for the H3.1-K27M cultures we measured an additive effect only in higher dosages of Vorinostat (Fig. 2C-E and S2E-F). In line with the difference in sensitivity, H3.1- and H3.3-K27M cells differed substantially in MYC expression levels, with H3.1-K27M cells expressing significantly lower levels of MYC (Fig. S2G). Indeed, BLISS indexes negatively correlated with MYC expression levels (Fig. 2F), indicating that H3.3-K27M cells, which express higher levels of MYC, are more sensitive to the combination of Sulfopin and Vorinostat. Importantly, this highlights MYC, and MYC target genes, as potential biomarkers of response for the combined treatment. For example, the expression levels of the two MYC targets, NOC4L and PSMD3, negatively correlated with the BLISS indexes and significantly differed between H3.1- and H3.3-K27M cells (Fig. 2F and S2H). Furthermore, these and additional MYC targets were downregulated following the combined treatment in the H3.3-K27M culture, SU-DIPG13 (Fig. S2I). Overall, our data demonstrates that combining Sulfopin treatment with Vorinostat results in an additive effect on the viability of DMG cultures, with H3.3-K27M cells showing higher sensitivity. This is in line with previous studies highlighting clinical and molecular distinction between H3.1- and H3.3-K27M mutant tumors, which is thought to dictate their targetable vulnerabilities5,47,48.

Combination of Sulfopin and Vorinostat elicits robust downregulation of oncogenic pathways

(A) Timeline demonstrating the treatment protocol for the combination of Sulfopin and Vorinostat.

(B) Percentage of cell viability, as measured by CellTiterGlo of SU-DIPG13 cells treated with Sulfopin and Vorinostat at the indicated concentrations, compared to DMSO.

(C) BLISS index measured as the ratio between the observed and the expected effect of the combination of Sulfopin and Vorinostat, for each pair of concentrations, in SU-DIPG13. Synergy: Bliss <1 ,Additive: Bliss=1, Antagonist: Bliss>1.

(D) Cell viability as measured by CellTiterGlo, of eight DMG cultures treated with Sulfopin (0uM, 2.5uM, 5uM, 10uM, 20uM and 40uM) and Vorinostat (1uM), compared to DMSO. H3.3-K27M and H3.1-K27M cultures are indicated in blue and orange, respectively. Mean±SD of two technical replicates is shown. H3.3-K27M cells showed higher sensitivity to the combined treatment compared to H3.1-K27M cells.

(E) The BLISS index of the combination of Sulfopin (10uM) and Vorinostat (1uM), in the indicated cultures. An additive effect was detected in all the H3.3-K27M cultures at this set of concentrations.

(F) Pearson correlation coefficient matrix of BLISS index of the combined treatment (Sulfopin (10uM) and Vorinostat (1uM)) and mRNA levels of MYC and its target genes, in the eight DMG cultures tested. mRNA levels were measured by RT-qPCR (Fig. S2F-G). Blue and yellow colors indicate negative or positive correlation, respectively. Negative correlation was detected between the BLISS indexes and the expression levels of MYC and its target genes.

(G) Unsupervised hierarchical clustering of expression levels of 620 significantly DE genes detected in SU-DIPG13 cells treated with either Sulfopin (10uM, 8 days), Vorinostat (1uM, 72 hours), the combination of Sulfopin and Vorinostat or DMSO. Gene expression rld values (log2 transformed and normalized) were standardized for each gene (row) across all samples. Color intensity corresponds to the standardized expression, low (blue) to high (red). Clusters 1 and 4 demonstrate additive transcriptional patterns associated with the combined treatment.

(H) Top: Gene Set Enrichment Analysis (GSEA) on SU-DIPG13 treated with combination of 10uM Sulfopin and 1uM Vorinostat compared to DMSO, showing significant downregulation of mTORC1 signaling (‘HALLMARK_MTORC1_SIGNALING’ geneset119) in the combined treatment. NES: Normalized Enrichment Score. FDR: false discovery rate. Bottom: Expression levels of significantly DE genes detected in the combined treatment compared to DMSO that are part of the mTORC1 signaling geneset. MTOR gene was added manually to the heat-map. Heatmaps were generated as described in B.

(I) Top: Gene Set Enrichment Analysis (GSEA) on SU-DIPG13 treated as in C, showing significant downregulation of the epigenetic BMI-1 pathway and the oncogenic cAMP pathway in the combined treatment (BMI1_DN.V1_UP; CAMP_UP.V1_UP; MSigDB C6 oncogenic signature120,121). Bottom: Expression levels of significantly DE genes detected in the combined treatment compared to DMSO that are part of the BMI-1 and cAMP genesets. Heatmaps were generated as in described B.

(J) Western blot of SU-DIPG13 treated either with Sulfopin (10uM, 8 days), Vorinostat (1uM, 72 hours), the combination of Sulfopin and Vorinostat or DMSO, using the indicated antibodies. β-tubulin is used as loading control.

To gain more insights on the molecular mechanisms underlying this effect on viability, we profiled the transcriptome of SU-DIPG13 cells (H3.3-K27M) treated with each of the drugs alone and the combined treatment (Fig. S2J). A total of 494 genes were significantly differentially expressed (DE) in cells treated with the drug combination, compared to only 145 and 167 in cells treated solely with Sulfopin or Vorinostat, respectively (Fig. S2K-L). Hierarchical clustering of all 620 DE genes confirmed strong additive effect of the combined treatment on gene expression, primarily demonstrated by clusters 1 and 4 (Fig. 2G). Cluster 1, consisting of approximately half of the DE genes, was strongly downregulated by the combined treatment, compared to single-agent treated cells. This cluster included genes that play crucial roles in glioma progression such as MTOR, ATF3, CREB3 and TGFA49–52 (Fig. S2M). Moreover, the mTORC1 signaling pathway was highly enriched within the genes comprising cluster 1 (Fig. S2N). Importantly, GSEA analysis revealed that this pathway was significantly enriched in downregulated genes following the combined treatment (Fig. 2H and S2O). Of note, additional epigenetic and oncogenic pathways such as BMI1, cAMP and IL-6/JAK/STAT3 were significantly downregulated in cells treated with the drug combination53–55 (Fig. 2I and S2O). Reduction in mTOR signaling (mTOR, phospho-mTOR and phospho-S656) and ATF3 was further validated by western blot and RT-qPCR analysis (Fig. 2J, S2P). Importantly, we also confirmed that the combined treatment with Sulfopin and Vorinostat led to induction of p21 (CDKN1A), a prominent tumor suppressor promoting cell cycle arrest and regulated by mTOR signaling57–59 (Fig. 2H, J). Finally, exploring mTOR expression in all the tested DMG cultures, revealed a strong negative correlation with the BLISS indexes of the combined treatment (Fig. S2Q), suggesting mTOR as a potential biomarker of response for the combined treatment of Sulfopin and Vorinostat. Overall, our data revealed that the combined inhibition of HDACs and MYC results in robust downregulation of prominent oncogenic pathways.

Sulfopin attenuates HDACi-mediated accumulation of H3K27ac on genomic regions associated with oncogenic pathways

We next examined epigenetic features that may underly the inhibition of oncogenic pathways by the combined treatment (keeping the same experimental set-up as shown in figure 2A). Inhibition of histone deacetylases is expected to increase and alter genome-wide histone acetylation patterns, including the H3K27ac mark associated with active promoters and enhancers60. Indeed, global quantification by single-molecule measurements of H3K27-acetylated nucleosomes, as well as western blot analysis, confirmed HDACi-mediated increase in H3K27ac global levels (Fig. 3A-C, S3A). Interestingly, treatment with Sulfopin alone had the opposite effect, reducing the percentage of acetylated nucleosomes (Fig. 3B-C, S3A). As acetylation is coupled with transcription activation, this is in line with the function of Sulfopin as an inhibitor of MYC transcriptional activity26,61. As a result, the global HDACi-mediated increase in histone acetylation was also attenuated in the combined treatment with Sulfopin (Fig. 3B-C, S3A).

The combined treatment attenuates H3K27ac levels on oncogenic targets

(A) Scheme of the single-molecule imaging experimental setup133: cell-derived mono-nucleosomes are anchored in a spatially distributed manner on polyethylene glycol (PEG)-coated surface. Captured nucleosomes are incubated with fluorescently labeled antibodies directed against the H3K27ac modification. Total internal reflection fluorescence (TIRF) microscopy is utilized to record the position and modification state of each nucleosome. Time series images are taken to allow detection of maximal binding events.

(B) Single-molecule imaging quantification of the percentage of H3K27ac nucleosomes, in SU-DIPG13 cells treated with either Sulfopin (10uM, 8 days), Vorinostat (1uM, 72 hours), or the combination of Sulfopin and Vorinostat, normalized to DMSO. Mean fold ± SE of at least two independent experiments is shown. H3K27ac global levels are lower in the combined treatment compared to cells treated solely with Vorinostat. *P < 0.05 (two sample t-test).

(C) SU-DIPG13 cells were treated as in B, and analyzed by western blot using the indicated antibodies.

(D) Left panel: Heatmap shows H3K27ac read coverage around the TSS (+/-5Kb) of the significantly DE genes shown in figure 2B, in SU-DIPG13 cells treated with the combination of 10uM Sulfopin and 1uM Vorinostat versus DMSO. Average coverage is shown on top. Color intensity corresponds to the standardized expression. Clusters 1-4 are indicated. Right panel: The log2 ratio of H3K27ac read coverage in SU-DIPG13 cells treated with the combination of 10uM Sulfopin and 1uM Vorinostat vs. DMSO was calculated. Heatmap shows the ratio around the TSS (+/-5Kb) of the significantly DE genes shown in figure 2B, and average coverage is shown on top. Color intensity corresponds to the ratio between samples, low (red) to high (blue). Clusters 1-4 are indicated, with cluster 1 presenting the strongest local decrease in H3K27ac following the combined treatment compared to DMSO.

(E) IGV tracks of MTOR and SLC7A5 gene promoters, showing H3K27ac coverage in SU-DIPG13 cells treated as indicated.

(F) Functional enrichment analysis of the genes linked to enhancers (top targets of high confident enhancers) marked with H3K27ac exclusively in SU-DIPG13 cells treated with Vorinostat, and not in the combined treatment. gProfiler algorithm126 was used to calculate enrichment against the KEGG pathways DB128. Dashed line denotes adjusted p-value = 0.05. Genes associated with Vorinostat-unique enhancers are enriched for oncogenic signaling pathways.

(G) IGV track of AKT3 and SEC13 linked enhancers, showing H3K27ac coverage in SU-DIPG13 cells treated with 1uM Vorinostat or the combination of 10uM Sulfopin and 1uM Vorinostat.

(H) Normalized expression levels of AKT3 and SEC13 genes in SU-DIPG13 cells treated as in G. Mean ± SD of three technical repeats is shown. *P < 0.05 (two-sample t-test).

To profile changes in the genomic distribution of H3K27ac and H3K27me3, we applied Cut&Run in SU-DIPG13 cells treated with Vorinostat alone or in combination with Sulfopin. We first examined the promoters of genes differentially expressed in the combined treatment (as clustered in Fig. 2G). Genes comprising cluster 1 showed higher basal expression levels compared to genes from cluster 4, and these expression differences were also reflected in higher H3K27ac at these genes’ promoters (Fig. S3B-C). Moreover, in accordance with the reduction in RNA expression levels of genes comprising cluster 1, the combined treatment led to the strongest local reduction of H3K27ac levels associated with their promoters (Fig. 3D). Specifically, we found that several of the oncogenes that comprise cluster 1, such as mTOR, and mTOR regulator-SLC7A5, RELB and COPB2, showed decreased levels of H3K27ac at their promoters following the combined treatment62–64 (Fig. 3E, S3D). Cluster 1 genes also gained H3K27me3 on their promoters upon the combined treatment, supporting their epigenetic repression following Sulfopin and Vorinostat treatment (Fig. S3E).

We further identified genomic regions in which histone acetylation is lost in a Sulfopin-dependent manner (i.e., H3K27ac peaks that are present in Vorinostat-treated cells, but are lost in the combined treatment with Sulfopin). Genome distribution analysis of these unique peaks revealed that the majority of them localize to distal regions (Fig. S3F) and more than 60% of them localize to annotated enhancers. Importantly, genes associated with these enhancers were strongly enriched for oncogenic pathways such as RAS, MAPK and PI3K-AKT signaling pathways66–68 (Fig. 3F, S3G). Specifically, the combined treatment reduced H3K27ac levels on several enhancers linked to genes involved in mTOR pathway and gliomagenesis, such as: AKT3, JAK1 and NFKB2 genes, that were shown to have critical role in malignant gliomas69–73, SEC13 gene that indirectly activates mTOR74,75, and CD93, a key regulator in glioma76 (Fig 3G and S3H). Concomitant with the lower H3K27ac levels associated with these enhancers, we also observed lower expression of the associated genes (Fig. 3H and S3I). Taken together, the results show attenuation of HDACi-mediated accumulation of H3K27ac levels upon combined treatment with Sulfopin, associated with promoters and enhancers of prominent oncogenes. This is in-line with reduced expression of these oncogenic pathways, strengthening the therapeutic potential of the combined treatment.

The combined treatment reduces tumor growth in a DMG xenograft mouse model

Finally, we examined whether the combined inhibition of HDACs and PIN1 elicits an additive effect also in-vivo. Thus, we established a H3-K27M DMG orthotopic xenograft model by injections of SU-DIPG13P* cells, engineered to express firefly luciferase, to the pons of immunodeficient mice17. Mice were given ten days to develop tumors, and then randomly assigned to four groups: control mice treated with DMSO, mice treated with Sulfopin or Vorinostat as single agents, and mice treated with the combination of Sulfopin and Vorinostat. Both drugs were given daily and tumor growth was monitored by in in-vivo bioluminescence imaging. In-line with our in-vitro results, we found significant reduction in tumor growth following the combined treatment compared to control DMSO treated mice (Fig. 4A-B). Importantly, the reduction was also seen when comparing to mice treated solely with Vorinostat, supporting the additive value of Sulfopin to the well-known anti-tumorigenic effect of HDACi in H3-K27M gliomas17 (Fig. 4A-B). Of note, mice from all groups showed signs of severe dehydration and general deterioration after 18 days of treatment, and thus we could not measure a survival benefit for the combined treatment. This rapid deterioration is likely a result of the aggressiveness of the transplanted tumors and does not represent side effects of the treatment, as mice from all groups, including the non-treated mice, showed similar signs of deterioration.

The combined treatment with Sulfopin and Vorinostat reduces tumor growth in-vivo

(A-B) SU-DIPG13P* cells were injected to the pons of immunodeficient mice to form tumors. Ten days post injection, mice were treated for 18 days with either DMSO, Sulfopin, Vorinostat, or the combination of Sulfopin and Vorinostat. (A) In-vivo bioluminescent imaging of DMG xenografts following 18 days of treatment. The heat map superimposed over the mouse head represents the degree of photon emission by DMG cells expressing firefly luciferase. (B) DMG xenograft tumor growth as measured by change in bioluminescent photon emission following 15 days of treatment with either DMSO (n=8), Sulfopin (n=4), Vorinostat (n=5) or the combination of Sulfopin and Vorinostat (n=6). Data points represent the fold-change in maximum photon flux between day 3 and day 18 under treatment for each mouse. *P < 0.05 (two-tailed Mann-Whitney U-test).

(C-D) Immunofluorescent staining of brain sections from mice treated with DMSO (n=4) or the combination of Sulfopin and Vorinostat (n=4). (C) Representative fluorescence images of H3-K27M (red) and mTOR (green). (D) Percentage of mTOR positive cells out of the total H3-K27M-positive cells (n=13-198 cells per FOV). H3-K27M positive cells show lower levels of mTOR following the combined treatment compared to DMSO. *P < 0.05 (two-tailed t-test).

To link our in-vitro transcriptomic analysis to the in-vivo effect on tumor growth, we assessed the percentage of H3-K27M cells that are positive to mTOR signal (double-positive), out of the total number of H3-K27M cells, by immunofluorescence staining. The combined treatment significantly reduced the fraction of the double-positive cells, compared to DMSO (Fig. 4C-D). Of note, mTOR signal in the mouse cells surrounding the tumor showed no significant differences between the DMSO and the combination-treated mice (Fig S4A-B). Overall, these results verify the potency of this treatment in restricting oncogenic pathways also in-vivo, and the potential clinical benefit of PIN1 and HDAC combined inhibition in H3-K27M-mutant DMGs.

Discussion

Despite decades of efforts to improve treatment for children diagnosed with H3-K27M-mutant DMGs, the standard of care remains solely radiation, with no significant improvement in overall survival1,2,79. High throughput drug screens in patient-derived cells identified HDAC inhibition as a promising therapeutic strategy, and ongoing clinical trials revealed preliminary evidence of clinical benefit38,80 (NCT02717455, NCT03566199). However, emerging resistance to HDACi in glioma cells17, as well as limited therapeutic benefits of HDACi treatment as a sole agent81,82, highlight the necessity to discover additional therapy modalities that could prolong survival and relieve symptoms.

In H3-K27M-mutant DMGs, as well as in many other malignancies, high MYC activity is correlated with poor prognosis and therefore considered as an attractive therapeutic target21,83. Nevertheless, direct targeting of MYC is highly challenging, functional and structural wise. MYC signaling, while amplified in cancer cells, is also essential to normal cell function84. Moreover, as a transcription factor, MYC lacks traditionally druggable binding pockets, hindering the development of direct small molecule binders25. To overcome this, in this study we exploited Sulfopin, a novel, potent and selective PIN1 inhibitor26. Pre-clinical studies have shown in-vivo activity and negligible toxicity of Sulfopin in murine models of PDAC and neuroblastoma26. While PIN1 is not overexpressed in DMG tumors, we show that Sulfopin, although not inhibiting MYC directly, downregulates MYC target genes in patient-derived DMG cells and affects cell survival in a H3-K27M-dependant manner. These results are in line with the contribution of MYC signaling to H3-K27M-driven tumorigenesis23,24, underscoring the therapeutic potential of Sulfopin for these aggressive brain tumors. Notably, despite a significant reduction in tumor size in-vivo, the combined treatment did not increase mice survival. This is perhaps due to the relatively large tumors already formed at the onset of treatment, leading to rapid deterioration of mice in all experimental groups. Thus, further optimization of the modeling system and therapeutic regime is needed.

We showed that combining Sulfopin treatment with the HDAC inhibitor Vorinostat reduced cell viability by 80% in several tumor-derived DMG lines originating from different patients. Importantly, the combination resulted in an additive effect not only on cell viability, but also on transcriptional programs. A key oncogenic pathway that was robustly downregulated in the combined treatment, in-vitro and in-vivo, is the mTOR signaling pathway. Preclinical and clinical data support a role for mTOR in gliomagenesis85. Specifically, AKT gain or PTEN loss is detected in approximately 70% of DMG tumors, and a recent study showed high expression of mTOR in DMGs. Thus, the PTEN/AKT/mTOR pathway is suggested to play a key role in this disease86–89. Attempts to restrict mTOR signaling using first generation mTOR inhibitors primarily inhibited mTORC1, which often caused upregulation of mTORC290. The combined treatment with Sulfopin and Vorinostat led to robust inhibition of several target proteins in the mTOR signaling pathway, likely via alternative pathways. One potential candidate is the mTOR-dependent cell-growth through MYC regulation, which may be impaired by Sulfopin treatment, inhibiting MYC transcriptional activity91–94. Of note, drug agents targeting both mTORC1 and mTORC2 were shown to be highly effective in DMG cells as well as in murine models95,96.

Expression of the H3-K27M oncohistone is associated with a prominent increase in H3K27ac levels11–13,16. Among the loci that accumulate this modification are repetitive elements and endogenous retroviruses (ERVs), resulting in their aberrant expression, potentially inducing innate immune responses12. However, H3-K27M cells do not show an increase in interferon stimulated genes (ISGs), preassembly due to high levels of MYC, which is potent antagonist of interferon responses12,97. Thus, dual targeting of both HDAC and MYC pathways has the potential of inducing robust expression of ISGs, thus promoting tumor cell death mediated by viral-mimicry98.

The high levels of H3K27ac in H3-K27M cells was also linked to activation of super-enhancers, which regulate the expression of oncogenes and genes that are associated with an undifferentiated state79. Here, we show that H3-K27M cells are vulnerable to transcriptional disruption mediated by the combined treatment with Sulfopin and Vorinostat. While treating cells with Vorinostat alone increased H3K27ac levels, as expected, the combined treatment attenuated the accumulation of this mark on promoters and enhancers of prominent oncogenes, resulting in their decreased expression. This Sulfopin-mediated attenuation could potentially result from downregulation of MYC targets, as it was recently shown that MYC overexpression results in H3K27ac gain on super-enhancers99.

In summary, we propose a novel therapeutic strategy targeting epigenetic and transcriptional pathways in H3-K27M-mutant DMG tumors. HDAC inhibition, in combination with the novel MYC inhibitor, inhibited oncogenic transcriptional signatures and reduced tumor growth in-vivo. Our findings emphasize epigenetic dependencies of H3-K27M driven gliomas and highlight the therapeutic potential of combined treatments for this aggressive malignancy.

Supporting information